Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-01-23 manuscript/manuscript.tex
edits to abstract

\newcommand\skhcomment[1]{{\color{cyan}#1}}
\newcommand\jdbcomment[1]{{\color{red}#1}}

@@

\title{Experimentally Informed Site-Specific Substitution Models Substantially Deepen Viral Divergence Estimates}

\author
{Sarah K. Hilton$^{1,2}$  and Jesse D. Bloom$^{1,2,}$\\
\\
\normalsize{$^1$Division of Basic Sciences and Computational Biology Program,}\\
\normalsize{Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}\\
\normalsize{$^2$Department of Genome Sciences, University of Washington, Seattle, WA}\\
\normalsize{E-mail:  jbloom@fredhutch.org.}\\
}

@@

\begin{abstract}
\textit{$\leq$ 250 words, currently 164}

Molecular dating techniques are often used to estimate the origin date of modern, circulating viruses. 
These estimates are often substantially younger than estimates from methods which do not rely on viral phylogenies. 
This discrepancy is thought to be caused in part by inadequate modeling of purifying selection leading to branch length underestimation. 
Here, we show that substitution models informed by empirical measurements of mutational tolerance model constraint better than traditional models and extend branch lengths. 
We used models informed by deep mutational scanning experiments performed in two, highly diverged influenza virus hemagglutinin homologs to optimize the branch lengths of a phylogenetic tree. 
For each experimentally informed model, we observed extension in branch length from the experiment's focal sequence. 
This extension in branch length due to explicit modeling of site-specific purifying selection is observed in the presence and absence of standard methods for modeling site-to-site variation. 
Overall, this study underscores the importance of modeling purifying selection when estimating branch lengths and, by extension, divergence dates.

\end{abstract}

>>>>

\newcommand\skhcomment[1]{{\color{cyan}[#1]}}
\newcommand\jdbcomment[1]{{\color{red}[#1]}}

@@

\title{Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages}

\author
{Sarah K. Hilton$^{1,2}$  and Jesse D. Bloom$^{1,2}$\\
\\
\normalsize{$^1$Division of Basic Sciences and Computational Biology Program,}\\
\normalsize{Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}\\
\normalsize{$^2$Department of Genome Sciences, University of Washington, Seattle, WA}\\
\normalsize{E-mail:  jbloom@fredhutch.org.}\\
}

@@

\begin{abstract}
\noindent \skhcomment{$\leq$ 250 words, currently 164}   
Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that substitution models informed by experimental measurements of site-specific purifying selection lengthen the branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to better modeling how site-specific amino-acid preferences affect the stationary state of the substitution models, and is independent of the branch-length-extension due to modeling site-to-site variation in substitution rate.
We show that the deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state as well as the substitution rate when estimating deep divergence times. 
\end{abstract}
